Linear Regret Bounds for Polyhedral Surrogates

### 2021-05-20 notes/regret-bound.tex
lower bound

\begin{theorem}[new version; weaker assumptions]
  Let $L:\reals^d \to \reals^\Y_+$ be a surrogate such that for all $y\in\Y$, the function $L(\cdot)_y$ is differentiable with a locally Lipschitz gradient.
  Let $\ell:\R \to \reals^\Y_+$ be a target loss and $\psi:\reals^d\to\R$ a link.
  Let $\Gamma := \prop{L}$, $\gamma := \prop{\ell}$.
  Suppose that there exist reports $r,r'$, distribution $p_0 \in \gamma_r \cap \gamma_{r'}$, and surrogate report $u_0 \in \Gamma(p_0)$ such that $u \mapsto \inprod{p}{L(u)}$ is $\alpha$-strongly convex in an open neighborhood of $u_0$.
  Then for any surrogate regret bound between $L$ and $\ell$ with function $\zeta:\reals_+\to\reals_+$, there exists $c > 0$ such that for sufficiently small $\epsilon>0$ we have
  \begin{equation}
    \label{eq:regret-lower-bound}
    \zeta(\epsilon) \geq c \cdot \sqrt{\epsilon}~.
  \end{equation}
\end{theorem}
\begin{proof}
  Assume without loss of generality that $\psi(u_0) = r'$; otherwise, swap the roles of $r$ and $r'$.
  As $\ell$ is non-redundant by assumption, there exists some $p_1 \in \inter{\gamma_r}$.
  We therefore have $R_\ell(r',p_1) = \inprod{p_1}{\ell(r')-\ell(r)} =: c_\ell > 0$, and $R_\ell(r',p_0) = 0$.
  Let $p_\lambda := (1-\lambda) p_0 + \lambda p_1$.
  By convexity of $\gamma_r$, we have $p_\lambda \in \gamma_r$ for all $\lambda \in [0,1]$, which gives $R_\ell(r',p_\lambda) = \lambda c_\ell$.
  We will upper bound $R_L(u_0,p_\lambda)$ by $O(\lambda^2)$, which will imply the result.
  % Our goal is to find some $c>0$ and some path of reports $u_\lambda \in \reals^d$ such that $\psi(u_\lambda)=r'$ and $R_L(u_\lambda,p_\lambda) \geq c\cdot\sqrt{\lambda \epsilon_1}$, for $\lambda$ sufficiently small.

  Let $L_\lambda:\reals^d\to\reals_+$ be given by $L_\lambda(u) = \inprod{p_\lambda}{L(u)} = (1-\lambda) \inprod{p_0}{L(u)} + \lambda \inprod{p_1}{L(u)}$.
  Let $\delta >0$ such that the above open neighborhood of $u_0$ contains the ball $B_\delta(u_0)$ of radius $\delta$.
  Let $u_1 \in \Gamma(p_1)$.
  For any $u\notin B_\delta(u_0)$, we have
  \begin{align*}
    L_\lambda(u) - L_\lambda(u_0)
    &=
      (1-\lambda) \left( L_0(u) - L_0(u_0) \right)
      + \lambda \left( L_1(u) - L_1(u_0) \right)
    \\
    &\geq
      (1-\lambda) \left( \frac \alpha 2 \delta^2 \right)
      + \lambda \left( L_1(u_1) - L_1(u_0) \right)~,
  \end{align*}
  where we used local strong convexity and the optimality of $u_1$.
  Thus, for $\lambda \leq \lambda^*$, where
  $\lambda^* := \alpha \delta^2 / (\alpha \delta^2 + 2L_1(u_0) - 2L_1(u_1))$,
  we have that $L_\lambda$ is minimized within $B_\delta(u_0)$.
  By $\alpha$-strong convexity of $L_0$ on $B_\delta(u_0)$, $L_\lambda$ is therefore $\alpha^*$-strongly convex on $B_\delta(u_0)$ for $\lambda \leq \lambda^*$, where $\alpha^* = (1-\lambda^*)\alpha$.\raft{Need to double-check this statement}
  For the remainder of the proof, let $\lambda \leq \lambda^*$.

@@

  We now upper bound $\|u_\lambda - u_0\|_2$, and then apply strong smoothness \raft{last hole to fill: use \emph{locally} Lipschitz instead of smoothness; should be fine since we apply it in a bounded region, e.g. around $u_1$} to upper bound $L_\lambda(u_0) - L_\lambda(u_\lambda)$.
  Consider the first-order optimality condition of $L_\lambda$:
  \begin{align*}
    \label{eq:first-order-opt-smooth}
    & 0 = \nabla L_\lambda(u_\lambda) = (1-\lambda) \nabla L_0(u_\lambda) + \lambda \nabla L_1(u_\lambda)
    \\
    & \implies (1-\lambda) \|\nabla L_0(u_\lambda)\|_2 = \lambda \|\nabla L_1(u_\lambda)\|_2~.
  \end{align*}
  By optimality of $u_0$ and $u_1$, strong convexity and smoothness, and the triangle inequality, we have
  \begin{align*}
    \|\nabla L_0(u_\lambda)\|_2 &= \|\nabla L_0(u_\lambda) - \nabla L_0(u_0)\|_2 \geq \alpha \|u_\lambda - u_0\|_2~,
    \\
    \|\nabla L_1(u_\lambda)\|_2 &= \|\nabla L_0(u_\lambda) - \nabla L_1(u_1)\|_2 \leq \beta \|u_\lambda - u_1\|_2
    \\
    &\leq \beta \left( \|u_\lambda - u_0\|_2 + \|u_0 - u_1\|_2 \right)~.
  \end{align*}
  Combining, and taking $\lambda \leq \tfrac 1 2 \tfrac {\alpha}{\alpha+\beta}$, we have
  \begin{align*}
    \|u_\lambda - u_0\|_2 \leq \frac{\lambda\beta}{(1-\lambda)\alpha-\lambda\beta} \|u_0 - u_1\|_2  \leq \lambda \frac{2\beta}{\alpha} \|u_0 - u_1\|_2 ~.
  \end{align*}
  Finally, from strong smoothness and optimality of $u_\lambda$, we have
  \begin{align*}
    L_\lambda(u_0) - L_\lambda(u_\lambda) \leq \frac{\beta}{2} \|u_0 - u_\lambda\|_2^2 \leq \frac{\beta}{2} \left(\lambda \frac{2\beta}{\alpha} \|u_0 - u_1\|_2\right)^2 = c_L \lambda^2~,
  \end{align*}
  where $c_L = \frac{2\beta^3}{\alpha^2} \|u_0 - u_1\|_2^2$.
  Take $c = \frac{c_\ell}{\sqrt{c_L}} = \frac{c_\ell \alpha}{\|u_0 - u_1\|_2 \sqrt{2\beta^3}}$.
  Recall $R_\ell(r',p_\lambda) = c_\ell \lambda$ and $R_L(u_0,p_\lambda) \leq c_L \lambda^2$.
  Then letting $\epsilon = R_L(u_0,p_\lambda)$, we have $\zeta(\epsilon) \geq R_\ell(r',p_\lambda) \geq c \sqrt{\epsilon}$.
\end{proof}

>>>>

\begin{theorem}[new version; weaker assumptions]
  Let $L:\reals^d \to \reals^\Y_+$ be a surrogate such that for all $y\in\Y$, the function $L(\cdot)_y$ is differentiable with a locally Lipschitz gradient.
  Let $\ell:\R \to \reals^\Y_+$ be a target loss and $\psi:\reals^d\to\R$ a link.
  Let $\Gamma := \prop{L}$, $\gamma := \prop{\ell}$.
  Suppose that there exist reports $r,r'\in\R$, distribution $p_0 \in \gamma_r \cap \gamma_{r'}$, and surrogate report $u_0 \in \Gamma(p_0)$ such that $u \mapsto \inprod{p}{L(u)}$ is $\alpha$-strongly convex in an open neighborhood of $u_0$.
  Then for any surrogate regret bound between $L$ and $\ell$ with function $\zeta:\reals_+\to\reals_+$, there exists $c > 0$ such that for sufficiently small $\epsilon>0$ we have
  \begin{equation}
    \label{eq:regret-lower-bound}
    \zeta(\epsilon) \geq c \cdot \sqrt{\epsilon}~.
  \end{equation}
\end{theorem}
\begin{proof}
  Assume without loss of generality that $\psi(u_0) = r'$; otherwise, swap the roles of $r$ and $r'$.
  As $\ell$ is non-redundant by assumption, there exists some $p_1 \in \inter{\gamma_r}$.
  We therefore have $R_\ell(r',p_1) = \inprod{p_1}{\ell(r')-\ell(r)} =: c_\ell > 0$, and $R_\ell(r',p_0) = 0$.
  Let $p_\lambda := (1-\lambda) p_0 + \lambda p_1$.
  By convexity of $\gamma_r$, we have $p_\lambda \in \gamma_r$ for all $\lambda \in [0,1]$, which gives $R_\ell(r',p_\lambda) = \lambda c_\ell$.
  We will upper bound $R_L(u_0,p_\lambda)$ by $O(\lambda^2)$, which will imply the result.
  % Our goal is to find some $c>0$ and some path of reports $u_\lambda \in \reals^d$ such that $\psi(u_\lambda)=r'$ and $R_L(u_\lambda,p_\lambda) \geq c\cdot\sqrt{\lambda \epsilon_1}$, for $\lambda$ sufficiently small.

  Let $L_\lambda:\reals^d\to\reals_+$ be given by $L_\lambda(u) = \inprod{p_\lambda}{L(u)} = (1-\lambda) \inprod{p_0}{L(u)} + \lambda \inprod{p_1}{L(u)}$.
  Let $\delta >0$ such that the above open neighborhood of $u_0$ contains the Euclidean ball $B_\delta(u_0)$ of radius $\delta$ around $u_0$.
  Let $u_1 \in \Gamma(p_1)$.
  For any $u\notin B_\delta(u_0)$, we have
  \begin{align*}
    L_\lambda(u) - L_\lambda(u_0)
    &=
      (1-\lambda) \left( L_0(u) - L_0(u_0) \right)
      + \lambda \left( L_1(u) - L_1(u_0) \right)
    \\
    &\geq
      (1-\lambda) \left( \frac \alpha 2 \delta^2 \right)
      + \lambda \left( L_1(u_1) - L_1(u_0) \right)~,
  \end{align*}
  where we used local strong convexity and the optimality of $u_1$.
  Thus, for $\lambda \leq \lambda^*$, where
  $\lambda^* := \alpha \delta^2 / (2 \alpha \delta^2 + 4 L_1(u_0) - 4 L_1(u_1))$,
  we have that $L_\lambda$ is minimized within $B_\delta(u_0)$.
  By $\alpha$-strong convexity of $L_0$ on $B_\delta(u_0)$, $L_\lambda$ is therefore $\alpha^*$-strongly convex on $B_\delta(u_0)$ for $\lambda \leq \lambda^*$, where $\alpha^* = (1-\lambda^*)\alpha$.\raft{Need to double-check this statement; also need to propagate $\alpha^*$ where appropriate}
  For the remainder of the proof, let $\lambda \leq \lambda^*$.

@@

  By assumption, the gradient of $L_y$ is locally Lipschitz for all $y\in\Y$.
  We will apply this fact to the compact set $\mathcal C = \{u \in \reals^d : \|u - u_1\| \leq \|u_0 - u_1\| + \delta\}$.
  By compactness, we have a finite subcover of open neighborhoods; let $\beta$ be the minimum Lipschitz constant over this finite set of neighborhoods.
  We thus have that $L_y$ is $\beta$-strongly smooth on $\mathcal C$, and hence so is $L_\lambda$ for any $\lambda \in [0,1]$.

  We now upper bound $\|u_\lambda - u_0\|_2$, and then apply strong smoothness to upper bound $L_\lambda(u_0) - L_\lambda(u_\lambda)$.
  Consider the first-order optimality condition of $L_\lambda$:
  \begin{align*}
    \label{eq:first-order-opt-smooth}
    & 0 = \nabla L_\lambda(u_\lambda) = (1-\lambda) \nabla L_0(u_\lambda) + \lambda \nabla L_1(u_\lambda)
    \\
    & \implies (1-\lambda) \|\nabla L_0(u_\lambda)\|_2 = \lambda \|\nabla L_1(u_\lambda)\|_2~.
  \end{align*}
  By optimality of $u_0$ and $u_1$, strong convexity and smoothness, and the triangle inequality, we have
  \begin{align*}
    \|\nabla L_0(u_\lambda)\|_2 &= \|\nabla L_0(u_\lambda) - \nabla L_0(u_0)\|_2 \geq \alpha \|u_\lambda - u_0\|_2~,
    \\
    \|\nabla L_1(u_\lambda)\|_2 &= \|\nabla L_1(u_\lambda) - \nabla L_1(u_1)\|_2 \leq \beta \|u_\lambda - u_1\|_2
    \\
    &\leq \beta \left( \|u_\lambda - u_0\|_2 + \|u_0 - u_1\|_2 \right)~.
  \end{align*}
  Combining, we have
  \begin{align*}
    (1-\lambda) \alpha \|u_\lambda - u_0\|_2
    &\leq
      (1-\lambda) \|\nabla L_0(u_\lambda)\|_2
    \\
    &= \lambda \|\nabla L_1(u_\lambda)\|_2
    \\
    &\leq
      \lambda \beta \left( \|u_\lambda - u_0\|_2 + \|u_0 - u_1\|_2 \right)~.
  \end{align*}
  Now rearranging and taking $\lambda \leq \tfrac 1 2 \tfrac {\alpha}{\alpha+\beta}$, we have
  \begin{align*}
    \|u_\lambda - u_0\|_2 \leq \frac{\lambda\beta}{(1-\lambda)\alpha-\lambda\beta} \|u_0 - u_1\|_2  \leq \lambda \frac{2\beta}{\alpha} \|u_0 - u_1\|_2 ~.
  \end{align*}
  Finally, from strong smoothness and optimality of $u_\lambda$, we have
  \begin{align*}
    L_\lambda(u_0) - L_\lambda(u_\lambda) \leq \frac{\beta}{2} \|u_0 - u_\lambda\|_2^2 \leq \frac{\beta}{2} \left(\lambda \frac{2\beta}{\alpha} \|u_0 - u_1\|_2\right)^2 = c_L \lambda^2~,
  \end{align*}
  where $c_L = \frac{2\beta^3}{\alpha^2} \|u_0 - u_1\|_2^2$.
  Take $c = \frac{c_\ell}{\sqrt{c_L}} = \frac{c_\ell \alpha}{\|u_0 - u_1\|_2 \sqrt{2\beta^3}}$.
  Recall $R_\ell(r',p_\lambda) = c_\ell \lambda$ and $R_L(u_0,p_\lambda) \leq c_L \lambda^2$.
  Then letting $\epsilon = R_L(u_0,p_\lambda)$, we have $\zeta(\epsilon) \geq R_\ell(r',p_\lambda) \geq c \sqrt{\epsilon}$.
\end{proof}
